Implementing Equational Constraints\\ in a Functional Language

### 2011-07-12 papers/wlp2011/paper.tex

In this paper we consider KiCS2 \cite{BrasselFischerHanusReck11},
a new system that compiles functional logic programs of the
source language Curry \cite{Hanus06Curry}
into purely functional Haskell programs.
We have shown in \cite{BrasselFischerHanusReck11}
that this implementation can compete with or outperform
other existing implementations of Curry.
KiCS2 is based on the idea to represent
the search space, i.e., all non-deterministic results of a computation,
as a data structure that can be traversed by operations
implementing various strategies.
Furthermore, logic variables are replaced by generators,
i.e., operations that non-deterministically evaluate to
all possible ground values of the type of the logic variable.
It has been shown \cite{AntoyHanus06ICLP}
that computing with logic variables by narrowing \cite{Reddy85,Slagle74}
or computing with generators by rewriting
are equivalent, i.e., compute the same values.
Although this implementation technique is correct \cite{Brassel11Thesis},
the generation of all values for logic variables
might be inefficient for applications that
exploit constraints on partial values.
For instance, in Prolog the equality constraint \ccode{X=c(a)}
is solved by instantiating the variable \code{X} to \code{c(a)},
but the equality constraint \ccode{X=Y} is solved by binding \code{X}
to \code{Y} without enumerating any values for \code{X} or \code{Y}.
Therefore, we propose in this paper new techniques
to implement equational constraints in the framework of KiCS2
(note that, in contrast to Prolog, unification is performed
modulo function evaluation).
Furthermore, we also show how functional patterns \cite{AntoyHanus05LOPSTR},
i.e., patterns containing evaluable operations providing
for more powerful pattern matching than in logic or functional languages,
can be implemented in this framework.
We show that both extensions lead to efficiency improvements
without sacrificing the efficiency of the kernel implementation.

@@

If non-deterministic operations are used as arguments in other operations,
a semantical ambiguity might occur. Consider the operations
%
\begin{curry}
  xor True  False = True
  xor True  True  = False
  xor False x     = x

  xorSelf x = xor x x
\end{curry}
%
and the expression \ccode{xorSelf aBool}.
If we interpret this program as a term rewriting system,
we could have the reduction
\begin{haskell}
  xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool
                 $\to~$  xor True False      $\to~$  True
\end{haskell}
leading to the unintended result \code{True}.
Note that this result cannot be obtained if we use a strict strategy
where arguments are evaluated prior to the function calls.
In order to avoid dependencies on the evaluation strategies
and exclude such unintended results,
Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed
the rewriting logic CRWL as a logical
(execution- and strategy-independent) foundation for declarative
programming with non-strict and non-deterministic operations.  This
logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
\label{ctc-semantics}
where values of the arguments of an operation are determined before the
operation is evaluated. This can be enforced in a lazy strategy
by sharing actual arguments.
For instance, the expression above can be lazily evaluated
provided that all occurrences of \code{aBool}
are shared so that all of them reduce either to \code{True} or to \code{False}
consistently.

@@

To understand the extensions described in the subsequent sections,
we sketch the translation of Curry programs
into Haskell programs as performed by KiCS2.
More details about this translation scheme can be found in
\cite{BrasselFischer08IFL,BrasselFischerHanusReck11}.

@@

To show a concrete example, we use the following implementation
of \code{IDSupply} based on unbounded integers:
%
\begin{haskell}
  type IDSupply = Integer
  initSupply    = return 1
  thisID      n = n
  leftSupply  n = 2 * n
  rightSupply n = 2 * n + 1
\end{haskell}
%
If we apply the same transformation to the rules defining \code{xor}
(actually, \code{xor} performs pattern matching on two arguments
which is avoided in KiCS2 by introducing an intermediate operation,
see \cite{BrasselFischerHanusReck11}) and evaluate the main expression
\code{(main 1)},
we obtain the result
\begin{haskell}
  Choice 2 (Choice 2 False True) (Choice 2 True False)
\end{haskell}
Thus, the result is non-deterministic and contains several choices,
but all of them have the same identifier.
This identifier is used when the values contained in this result
are searched for: one has to make \emph{consistent} selections in choices with
same identifiers. Thus, if we select the left branch as the value
of the outermost \code{Choice}, we also have to select the left branch
in the selected argument \code{(Choice 2 False True)} so that only
the value \code{False} is possible here.
Similarly, if we select the right branch as the value of the outermost
\code{Choice}, we also have to select the right branch in
its selected argument \code{(Choice 2 True False)} which yields the only
value \code{False}.
Thus, the unintended value \code{True} is not produced.

In general, to extract all values from a \code{Choice} structure,
we have to traverse it, compute all possible choices
but consider the choice identifiers to make consistent (left/right)
decisions.
The latter requirement can be implemented by storing
the decisions already made for some choices during the traversal.
For this purpose, we introduce the type
\begin{haskell}
  data Decision = NoDecision | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoDecision} represents the fact that the value of a choice
is yet undecided.
Furthermore, we assume operations to lookup the current decision
for a given identifier or change its decision (depending on the implementation
of \code{IDSupply}, KiCS2 supports various implementations
based on memory cells or finite maps):
\begin{haskell}
  lookupDecision :: ID -> IO Decision
  setDecision    :: ID -> Decision -> IO ()
\end{haskell}
%
Now we can print all values contained in a choice structure
in a depth-first manner by the following I/O operation:\footnote{%
Note that this code has been simplified for readability
since the type system of Haskell does not
allow this direct definition.}
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: a -> IO ()
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
   where
    follow ChooseLeft  = printValsDFS x1
    follow ChooseRight = printValsDFS x2
    follow NoDecision  = do newDecision ChooseLeft  x1
                            newDecision ChooseRight x2

    newDecision d x = do setDecision i d
                         printValsDFS x
                         setDecision i NoDecision
  printValsDFS v                = print v
\end{haskell}
This operation ignores failures and prints values that are not a \code{Choice}.
For a \code{Choice}, it checks whether a decision for
this identifier has already been made (note that the initial value
for all identifiers is \code{NoDecision}).
If a decision has been made for this choice, it follows this decision.
Otherwise, the left alternative is used and this decision is stored.
After printing
all the values w.r.t.\ this decision,
the decision is undone (like in backtracking)
and the right alternative is used and stored.

@@

To avoid an unnecessary growth of the search space represented by
\code{Choice} constructors, our compiler performs an optimization for
deterministic operations. If an operation does not call, 
neither directly nor indirectly through
other operations, the choice operation \ccode{?},\footnote{%
Note that before applying this compilation scheme, overlapping rules in
user-defined operations are transformed into calls of the choice operation
\ccode{?}.}
then it is not necessary to pass an identifier supply to the operation.
In consequence, only the matching rules are extended so that the generated code
is nearly identical to a corresponding functional program.
Actually, the benchmarks presented in \cite{BrasselFischerHanusReck11}
show that this implementation outperforms all other Curry implementations
for deterministic operations,
and, for non-deterministic operations, outperforms Prolog-based
implementations of Curry and can compete with MCC \cite{Lux99FLOPS},
a Curry implementation that compiles to C.

As mentioned in the introduction,
KiCS2 translates occurrences of logic variables
into generators. For instance, consider the operation \code{not}
defined by the rules:
\begin{curry}
  not True  = False
  not False = True
\end{curry}
Then the expression \ccode{not x}, where \code{x} is a logic variable,
is translated into \ccode{not aBool} (we omit the \code{IDSupply} argument
here). The latter expression is evaluated by reducing the
argument \code{aBool} to a choice between \code{True} or \code{False}
followed by applying \code{not} to this choice.
This is similar to a narrowing step on \ccode{not x}
that instantiates the variable \code{x} to \code{True} or \code{False}.
% Actually, it has been shown in \cite{AntoyHanus06ICLP}
% that narrowing expressions with logic variables
% is equivalent to rewriting with generators.
Since such generators are standard
non-deterministic operations, they are translated like any other operation
so that one does not need additional run-time support for them.
However, if equational constraints should be solved,
there are methods which are more efficient than generating all values.
These methods and their implementation are discussed in the
next section.

@@

In this section we evaluate our implementation by some benchmarks.
They were executed on a Linux machine running Debian 5.0.7 with
an Intel Core 2 Duo (3.0GHz) processor.
KiCS2 has been executed with
the Glasgow Haskell Compiler (GHC 7.0.4, option -O2) and uses
an efficient \code{IDSupply} implementation that makes use of \code{IORef}s.
For a comparison with other mature implementations of Curry,
we considered PAKCS \cite{Hanus10PAKCS}
(Version 1.9.2, based on a SICStus-Prolog 4.1.2)
and MCC \cite{Lux99FLOPS} (Version 0.9.10). The timings were performed with the
time command measuring the execution time (in seconds) of a compiled
executable for each benchmark as a mean of three runs.
The programs used for the benchmarks,
partially taken from \cite{AntoyHanus05LOPSTR},
are \code{last} (compute the last element of a list),\footnote{%
\ccode{inc x n} is a naive addition that \code{n} times
increases its argument \code{x} by 1.}
\code{simplify} (simplify a symbolic arithmetic expression),
\code{varInExp} (non-deterministically return a variable occuring
                  in a symbolic arithmetic expression),
\code{half} (compute the half of a peano number using logic variables),
and
\code{palindrome} (check whether a list is a palindrome).

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & \code{==~}& \code{=:=} & \code{=:<=} \\
\hline
\code{last (map (inc 0) [1..100000])}   & 134.23    & 0.83      & 0.26        \\
% \code{grep}                             & 1.06      & 0.19      &             \\
\code{simplify}                         & 8.50      & 7.15      & 7.34        \\
\code{varInExp}                         & 11.72     & 1.50      & 1.40        \\
% \code{varInExp2}                        & 423.01    & 10.50     & 1.04        \\
% \code{varInExp3}                        & 399.68    & 16.70     & 11.28       \\
\code{fromPeano (half (toPeano 10000))} & 32.92     & 6.35      & 11.88       \\
\code{palindrome}                       & 28.59     & 15.58     & 23.21       \\
% \code{horseman}                         & 3.58      & 3.13      &             \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different equation solving methods}
 \label{fig:search-strategies}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & KiCS2 & PAKCS & MCC  \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.83  & 4.25   & 0.14 \\
% \code{grep}                             & 0.19  & 2.88   & 0.14 \\
\code{simplify}                         & 7.15  & 0.15   & 0.00 \\
\code{varInExp}                         & 1.50  & 3.89   & 0.30 \\
% \code{varInExp2}                        & 10.50 & 3.48   & stckOflw \\
% \code{varInExp3}                        & 16.70 & 19.22  & 1.18 \\
\code{fromPeano (half (toPeano 10000))} & 6.35  & 108.88 & 3.22 \\
\code{palindrome}                       & 15.58 & 32.56  & 1.07 \\
% \code{horseman}                         & 3.13  & 8.70   & 0.42 \\
\hline
\end{tabular}
\caption{Benchmarks: equational constraints in different Curry implementations}
 \label{fig:unification}
\end{figure}

@@

We have presented an implementation of equational constraints
and functional patterns in a purely functional implementation
of Curry. Our implementation is based on adding binding constraints
to computed values and processing them when values are extracted
at the top level of a computation.
Since we have added only new constructors and rules
for them in our implementation, it does not introduce
additional work if these features are not used, i.e.,
our implementation does not sacrifice the high efficiency
of the kernel implementation as shown in
\cite{BrasselFischerHanusReck11}.
However, if these features are used, they usually lead
to a more efficient execution, as demonstrated by our benchmarks.

>>>>

In this paper we consider KiCS2 \cite{BrasselHanusPeemoellerReck11},
a new system that compiles functional logic programs of the
source language Curry \cite{Hanus06Curry}
into purely functional Haskell programs.
We have shown in \cite{BrasselHanusPeemoellerReck11}
that this implementation can compete with or outperform
other existing implementations of Curry.
KiCS2 is based on the idea to represent
the search space, i.e., all non-deterministic results of a computation,
as a data structure that can be traversed by operations
implementing various strategies.
Furthermore, logic variables are replaced by generators,
i.e., operations that non-deterministically evaluate to
all possible ground values of the type of the logic variable.
It has been shown \cite{AntoyHanus06ICLP}
that computing with logic variables by narrowing \cite{Reddy85,Slagle74}
or computing with generators by rewriting
are equivalent, i.e., compute the same values.
Although this implementation technique is correct \cite{Brassel11Thesis},
the generation of all values for logic variables
might be inefficient for applications that
exploit constraints on partial values.
For instance, in Prolog the equality constraint \ccode{X=c(a)}
is solved by instantiating the variable \code{X} to \code{c(a)},
but the equality constraint \ccode{X=Y} is solved by binding \code{X}
to \code{Y} without enumerating any values for \code{X} or \code{Y}.
Therefore, we propose in this paper new techniques
to implement equational constraints in the framework of KiCS2
(note that, in contrast to Prolog, unification is performed
modulo function evaluation).
Furthermore, we also show how functional patterns \cite{AntoyHanus05LOPSTR},
i.e., patterns containing evaluable operations providing
for more powerful pattern matching than in logic or functional languages,
can be implemented in this framework.
We show that both extensions lead to efficiency improvements
without sacrificing the efficiency of the kernel implementation.

@@

If non-deterministic operations are used as arguments in other operations,
a semantical ambiguity might occur. Consider the operations
%
\begin{curry}
  xor True  x = not x             not True  = False
  xor False x = x                 not False = True

  xorSelf x = xor x x
\end{curry}
%
and the expression \ccode{xorSelf aBool}.
If we interpret this program as a term rewriting system,
we could have the reduction
\begin{haskell}
  xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool
                 $\to~$  xor True False      $\to~$  not False        $\to~$ True
\end{haskell}
leading to the unintended result \code{True}.
Note that this result cannot be obtained if we use a strict strategy
where arguments are evaluated prior to the function calls.
In order to avoid dependencies on the evaluation strategies
and exclude such unintended results,
Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed
the rewriting logic CRWL as a logical
(execution- and strategy-independent) foundation for declarative
programming with non-strict and non-deterministic operations.  This
logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
\label{ctc-semantics}
where values of the arguments of an operation are determined before the
operation is evaluated. This can be enforced in a lazy strategy
by sharing actual arguments.
For instance, the expression above can be lazily evaluated
provided that all occurrences of \code{aBool}
are shared so that all of them reduce either to \code{True} or to \code{False}
consistently.

@@

To understand the extensions described in the subsequent sections,
we sketch the translation of Curry programs
into Haskell programs as performed by KiCS2.
More details about this translation scheme can be found in
\cite{BrasselFischer08IFL,BrasselHanusPeemoellerReck11}.

@@

To show a concrete example, we use the following implementation
of \code{IDSupply} based on unbounded integers:
%
\begin{haskell}
  type IDSupply = Integer
  initSupply    = return 1
  thisID      n = n
  leftSupply  n = 2 * n
  rightSupply n = 2 * n + 1
\end{haskell}
%
If we apply the same transformation to the rules defining \code{xor}
(actually, \code{xor} performs pattern matching on two arguments
which is avoided in KiCS2 by introducing an intermediate operation,
see \cite{BrasselHanusPeemoellerReck11}) and evaluate the main expression
\code{(main 1)},
we obtain the result
\begin{haskell}
  Choice 2 (Choice 2 False True) (Choice 2 True False)
\end{haskell}
Thus, the result is non-deterministic and contains several choices,
but all of them have the same identifier.
This identifier is used when the values contained in this result
are searched for: one has to make \emph{consistent} selections in choices with
same identifiers. Thus, if we select the left branch as the value
of the outermost \code{Choice}, we also have to select the left branch
in the selected argument \code{(Choice 2 False True)} so that only
the value \code{False} is possible here.
Similarly, if we select the right branch as the value of the outermost
\code{Choice}, we also have to select the right branch in
its selected argument \code{(Choice 2 True False)} which yields the only
value \code{False}.
Thus, the unintended value \code{True} is not produced.

In general, to extract all values from a \code{Choice} structure,
we have to traverse it, compute all possible choices
but consider the choice identifiers to make consistent (left/right)
decisions.
The latter requirement can be implemented by storing
the decisions already made for some choices during the traversal.
For this purpose, we introduce the type
\begin{haskell}
  data Decision = NoDecision | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoDecision} represents the fact that the value of a choice
is yet undecided.
Furthermore, we assume operations to lookup the current decision
for a given identifier or change its decision (depending on the implementation
of \code{IDSupply}, KiCS2 supports various implementations
based on memory cells or finite maps):
\begin{haskell}
  lookupDecision :: ID -> IO Decision
  setDecision    :: ID -> Decision -> IO ()
\end{haskell}
%
Now we can print all values contained in a choice structure
in a depth-first manner by the following I/O operation:\footnote{%
Note that this code has been simplified for readability
since the type system of Haskell does not
allow this direct definition.}
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: a -> IO ()
  printValsDFS Fail             = return ()$\listline$
  printValsDFS (Choice i x1 x2) = lookupDecision i >>= follow
   where
    follow ChooseLeft  = printValsDFS x1
    follow ChooseRight = printValsDFS x2
    follow NoDecision  = do newDecision ChooseLeft  x1
                            newDecision ChooseRight x2$\listline$
    newDecision d x = do setDecision i d
                         printValsDFS x
                         setDecision i NoDecision$\listline$
  printValsDFS v = print v
\end{haskell}
This operation ignores failures and prints values that are not a \code{Choice}.
For a \code{Choice}, it checks whether a decision for
this identifier has already been made (note that the initial value
for all identifiers is \code{NoDecision}).
If a decision has been made for this choice, it follows this decision.
Otherwise, the left alternative is used and this decision is stored.
After printing
all the values w.r.t.\ this decision,
the decision is undone (like in backtracking)
and the right alternative is used and stored.

@@

To avoid an unnecessary growth of the search space represented by
\code{Choice} constructors, our compiler performs an optimization for
deterministic operations. If an operation does not call, 
neither directly nor indirectly through
other operations, the choice operation \ccode{?},\footnote{%
Note that before applying this compilation scheme, overlapping rules in
user-defined operations are transformed into calls of the choice operation
\ccode{?}.}
then it is not necessary to pass an identifier supply to the operation.
In consequence, only the matching rules are extended so that the generated code
is nearly identical to a corresponding functional program.
Actually, the benchmarks presented in \cite{BrasselHanusPeemoellerReck11}
show that this implementation outperforms all other Curry implementations
for deterministic operations,
and, for non-deterministic operations, outperforms Prolog-based
implementations of Curry and can compete with MCC \cite{Lux99FLOPS},
a Curry implementation that compiles to C.

As mentioned in the introduction,
KiCS2 translates occurrences of logic variables
into generators.
For instance, the expression \ccode{not x}, where \code{x} is a logic variable,
is translated into \ccode{not aBool} (we omit the \code{IDSupply} argument
here). The latter expression is evaluated by reducing the
argument \code{aBool} to a choice between \code{True} or \code{False}
followed by applying \code{not} to this choice.
This is similar to a narrowing step on \ccode{not x}
that instantiates the variable \code{x} to \code{True} or \code{False}.
% Actually, it has been shown in \cite{AntoyHanus06ICLP}
% that narrowing expressions with logic variables
% is equivalent to rewriting with generators.
Since such generators are standard
non-deterministic operations, they are translated like any other operation
so that one does not need additional run-time support for them.
However, if equational constraints should be solved,
there are methods which are more efficient than generating all values.
These methods and their implementation are discussed in the
next section.

@@

In this section we evaluate our implementation by some benchmarks.
They were executed on a Linux machine running Debian 5.0.7 with
an Intel Core 2 Duo (3.0GHz) processor.
KiCS2 has been executed with
the Glasgow Haskell Compiler (GHC 7.0.4, option -O2) and uses
an efficient \code{IDSupply} implementation that makes use of \code{IORef}s.
For a comparison with other mature implementations of Curry,
we considered PAKCS \cite{Hanus10PAKCS}
(Version 1.9.2, based on a SICStus-Prolog 4.1.2)
and MCC \cite{Lux99FLOPS} (Version 0.9.10). The timings were performed with the
time command measuring the execution time (in seconds) of a compiled
executable for each benchmark as a mean of three runs.
The programs used for the benchmarks,
partially taken from \cite{AntoyHanus05LOPSTR},
are \code{last} (compute the last element of a list),\footnote{%
\ccode{inc x n} is a naive addition that \code{n} times
increases its argument \code{x} by 1.}
\code{simplify} (simplify a symbolic arithmetic expression),
\code{varInExp} (non-deterministically return a variable occuring
                  in a symbolic arithmetic expression),
\code{half} (compute the half of a Peano number using logic variables),
\code{palindrome} (check whether a list is a palindrome),
\code{horseman} (solving an equation relating heads and feet of horses and men
based on Peano numbers),
and
\code{grep} (string matching based on a non-deterministic
specification of regular expressions \cite{AntoyHanus10CACM}).

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & \code{==~}& \code{=:=}& \code{=:<=} \\
\hline
\code{last (map (inc 0) [1..100000])}   & 134.23    & 0.83      & 0.26        \\
\code{simplify}                         & 8.50      & 7.15      & 7.34        \\
\code{varInExp}                         & 11.72     & 1.50      & 1.40        \\
% \code{varInExp2}                        & 423.01    & 10.50     & 1.04        \\
% \code{varInExp3}                        & 399.68    & 16.70     & 11.28       \\
\code{fromPeano (half (toPeano 10000))} & 32.92     & 6.35      & 11.88       \\
\code{palindrome}                       & 28.59     & 15.58     & 23.21       \\
\code{horseman}                         & 3.58      & 3.13      & n/a         \\
\code{grep}                             & 1.06      & 0.19      & n/a         \\
\hline
\end{tabular}
\caption{Benchmarks: comparing different equation solving methods}
 \label{fig:search-strategies}
\end{figure}

@@

\begin{figure}[t]
\centering
\begin{tabular}{|l|r|r|r|}
\hline
Expression                              & KiCS2 & PAKCS  & MCC  \\
\hline
\code{last (map (inc 0) [1..100000])}   & 0.83  & 4.25   & 0.14 \\
\code{simplify}                         & 7.15  & 0.15   & 0.00 \\
\code{varInExp}                         & 1.50  & 3.89   & 0.30 \\
% \code{varInExp2}                        & 10.50 & 3.48   & stckOflw \\
% \code{varInExp3}                        & 16.70 & 19.22  & 1.18 \\
\code{fromPeano (half (toPeano 10000))} & 6.35  & 108.88 & 3.22 \\
\code{palindrome}                       & 15.58 & 32.56  & 1.07 \\
\code{horseman}                         & 3.13  & 8.70   & 0.42 \\
\code{grep}                             & 0.19  & 2.88   & 0.14 \\
\hline
\end{tabular}
\caption{Benchmarks: equational constraints in different Curry implementations}
 \label{fig:unification}
\end{figure}

@@

We have presented an implementation of equational constraints
and functional patterns in a purely functional implementation
of Curry. Our implementation is based on adding binding constraints
to computed values and processing them when values are extracted
at the top level of a computation.
Since we have added only new constructors and rules
for them in our implementation, it does not introduce
additional work if these features are not used, i.e.,
our implementation does not sacrifice the high efficiency
of the kernel implementation as shown in
\cite{BrasselHanusPeemoellerReck11}.
However, if these features are used, they usually lead
to a more efficient execution, as demonstrated by our benchmarks.
